---
title: "Stat 579 - Life Expectancy"
author: "Joseph Zemmels"
date: "8/30/2018"
output: html_document
---
---
title: "Stat 579 - Life Expectancy"
author: "Joseph Zemmels"
date: "8/30/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=4, fig.height=4)
library(tidyverse)
```

# Life Expectancy in Mexico, Canada and the US 

## Step-by-step plan of work

1. Read the data
2. Focus on values of Canada
3. Visualize the values for Canada
4. See whether we can find the problem
5. ... then we'll see ...

## Reading the data 

```{r}
url <- "https://raw.githubusercontent.com/Stat579-at-ISU/stat579-at-isu.github.io/master/homework/data/gapminder-5060.csv" # nasty looking, but doesn't make any assumptions about data on your machine
life5060 <- read.csv(url)
head(life5060)
```

## Focus on the values for Canada

```{r}
canada <- life5060 %>% filter(country == "Canada")
head(canada)
```

## Visualize 

Draw a line for the life expectancy in Canada:
```{r}
canada %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_line()
```

Clearly there is something wrong with the value in 1957.

## Find the problematic value and fix it
#### **Note:** I wanted to practice piping, so I decided to replace the problematic value with the mean of the 2 surrounding values from '52 and '62.
```{r}
canada %>% filter(year == 1957)

lifeExp_1952 <- canada %>% filter(year == 1952) %>% .$lifeExp
lifeExp_1962 <- canada %>% filter(year == 1962) %>% .$lifeExp

canada_fixed <- canada %>% mutate(
  lifeExp = replace(lifeExp, year==1957, mean(c(lifeExp_1952,lifeExp_1962)))
)
```

## visualize the fixed data

```{r}
canada_fixed %>% ggplot(aes(x = year, y = lifeExp)) +
  geom_line()
```

## ... on to the stretch goal ... 

Add lines for Mexico and the US

```{r}
canada_fixed %>% 
  ggplot(aes(x = year, y = lifeExp, group=country)) + geom_line() +
  geom_line(data = life5060 %>% 
              filter(country %in% c("United States", "Mexico")))
```

## ... on to the stretch goal ... 

Add lines for Mexico and the US, color lines by country and add a legend

```{r, fig.width = 6}
canada_fixed %>% 
  ggplot(aes(x = year, y = lifeExp, colour=country)) + geom_line() +
  geom_line(data = life5060 %>% 
              filter(country %in% c("United States", "Mexico")))

```

## Expanded set with '70-'80 data included

```{r}
life7080 <- read.csv('gap7080.csv')

northAm5080 <- life5060 %>% rbind(life7080) %>% filter(country %in% c('United States','Canada','Mexico')) %>% arrange(country,year)

lifeExp_1952 <- northAm5080 %>% filter(country == 'Canada' & year == 1952) %>% .$lifeExp
lifeExp_1962 <- northAm5080 %>% filter(country == 'Canada' & year == 1962) %>% .$lifeExp

northAm5080_fixed <- northAm5080 %>% mutate(
  lifeExp = replace(lifeExp, country == 'Canada' & year==1957, mean(c(lifeExp_1952,lifeExp_1962)))
)

northAm5080_fixed
```

```{r}
northAm5080_fixed %>% ggplot(aes(x=year,y=lifeExp, color=country)) + geom_line() + labs(x='Years',y='Life Expectancy')
```

#### I notice that the life expectancy in Mexico changed much more drastically from 1952-1987 than in the U.S. and Canada. However, all 3 have been steadily increasing and the U.S. and Canada still have a decade-longer life expectancy.